Hypotheses

### 2025-07-02 analysis/data-exploration.qmd
Update
Updating analysis scripts

# Setting up {.unnumbered}

## Loading packages {.unnumbered}

```{r packages}
#| code-fold: true
#| code-summary: "Check our my packages"
#| warning: false
#| message: false

library(dplyr)
library(tidyr)
library(ggplot2)
library(ggalluvial)
```

## Loading functions {.unnumbered}

```{r functions}
#| code-fold: true
#| code-summary: "Check our my functions"

count_transitions <- function(data, years, absorbing = FALSE) {
  # Counts the occurrences of cognitive status transitions across specified 
  # years.
  # Converts numeric cognitive status codes (1, 2, 3) into descriptive labels
  # ("Normal Cognition", "MCI", "Dementia") and handles missing or unexpected values.
  # Aggregates the data to count transitions and reshapes it into a long format for analysis.
  # Arguments:
  #    - data: The input dataset containing cognitive function data.
  #    - years: A vector of years for which transitions should be counted.
  # Returns:
  #   - A dataset with counts of cognitive status transitions, including
  #   "Missing" and "Other" categories.

  # Create dynamic column names based on the years provided
  cogfunction_cols <- paste0("cogfunction", years)

  data <- data |>
    select(ID, all_of(cogfunction_cols)) |>
    na.omit() |>
    tidyr::pivot_longer(
      cols = !ID,
      names_to = "Wave",
      values_to = "Classification"
    ) |>
    mutate(
      Wave = stringr::str_replace(Wave, "cogfunction", "HRS "),
      Wave = factor(Wave, levels = c(paste0("HRS ", years))),

      Classification = case_when(
        Classification == 1 ~ "Normal Cognition",
        Classification == 2 ~ "MCI",
        Classification == 3 ~ "Dementia")
    )

  if(absorbing == TRUE){
    data <- data |>
      group_by(ID) |>
      mutate(
        Classification = ifelse(
          cumany(Classification == "Dementia"), "Dementia", Classification)) |>
      ungroup()
  }

  data <- data |>
    mutate(Classification = factor(
      Classification,
      levels = c("Normal Cognition", "MCI", "Dementia")))

  return(data)
}

plot_cognitive_scores <- function(data, year) {
  # Plots the distribution of cognitive scores over a range of years.
  # Converts the data from wide to long format and generates histograms for each wave.
  # Arguments:
  #   - data: The dataset containing cognitive scores.
  #   - year: The starting year for the range of cognitive scores to plot.
  # Returns:
  #   - A ggplot object showing the distribution of cognitive scores across waves.
  # Plotting dementia transitions
  data |>
    select(ID, any_of(paste0("cogtot27_imp", year:2022))) |>
    na.omit() |>
    tidyr::pivot_longer(cols = !ID,
                        names_to = "Wave",
                        values_to = "Score") |>
    mutate(Wave = as.double(stringr::str_replace(Wave, "cogtot27_imp", ""))) |>
    ggplot(aes(x = Score)) +
    geom_histogram(fill = "#56B4E9",
                   colour = "black",
                   alpha = 0.5) +
    scale_x_continuous(breaks = seq(0, 27, by = 3)) +
    labs(
      title = paste0("Distribution of cognitive scores ", year, "- 2022"),
      x = "Cognitive Score", y = "") +
    facet_wrap( ~ Wave)
}

plot_transitions <- function(data, size) {
  # Visualizes cognitive status transitions over time using an alluvial plot.
  # Represents the flow of individuals between cognitive states across waves.
  # Arguments:
  #   - data: The dataset containing transition data.
  #   - size: The font size for labels in the plot.
  # Returns:
  #   - A ggplot object showing transitions between cognitive states over time.

  require(ggalluvial)

  data |>
    ggplot(aes(x = Wave, y = n, stratum = Classification, 
               fill = Classification, alluvium = ID
    )) +
    geom_stratum(alpha = 0.5, width = 0.5) +
    geom_flow(width = 0.5) +
    geom_text(
      stat = "stratum",
      aes(label = stringr::str_wrap(Classification, width = 10)),
      hjust = 0.5,
      vjust = 0.5,
      size = size
    ) +
    labs(title = "Dementia transitions across time", 
         x = "", y = "Frequency") +
    scale_fill_viridis_d(direction = -1) +
    theme(axis.text = element_text(size = 10), 
          text = element_text(size = 10)) +
    ggeasy::easy_remove_legend()
}
```

## Setting theme {.unnumbered}

```{r theme}
#| code-fold: true
#| code-summary: "Check our my theme"

colour = "#212427"

theme_set(
theme_minimal() +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold", colour = colour),
    plot.subtitle = element_text(hjust = 0.5, size = 10, colour = colour),
    axis.title = element_text(size = 10, face = "bold", colour = colour),
    strip.text = element_text(size = 10, face = "bold", colour = colour)
    ))
```

## Reading in data {.unnumbered}

```{r data}
data <- read.csv(here::here("analysis/data/data.csv")) 
```

@@

```{r count-transitions}
dementia_data <- data |>
  count_transitions(years = seq(1996, 2022, by = 2))

@@

```{r missing-data}
#| warning: false
#| fig-width: 8
#| label: fig-missing-data
#| fig-cap: Occurance of missing data across HRS waves (participants were gathered from the 2020 HRS wave [in red])

@@

data |>
  select(any_of(paste0("cogtot27_imp", 1996:2022))) |>
  rename_with( ~ stringr::str_replace(., "cogtot27_imp", ""), starts_with("cogtot27")) |>
  visdat::vis_dat() +
  labs(title = "Missing data across HRS Waves (1996 - 2022)",
       subtitle = "Total Cognitive Scores") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold", colour = colour),
    plot.subtitle = element_text(hjust = 0.5, size = 12, face = "bold", colour = colour),
    panel.grid = element_blank(),
    axis.text.x = element_text(colour = colours)) +
  ggeasy::easy_remove_legend() +
  ggeasy::easy_x_axis_labels_size(size = 9)

@@

```{r complete-data}
#| fig-width: 8
#| label: fig-complete-data
#| fig-cap: HRS dataset after processing

data |>
  select(any_of(paste0("cogtot27_imp", 2016:2022))) |>
  rename_with( ~ stringr::str_replace(., "cogtot27_imp", ""), starts_with("cogtot27")) |>
  na.omit() |>
  visdat::vis_dat() +
  labs(
    title = "Complete dataset across HRS Waves (2016 - 2022)",
    subtitle = "Total Cognitive Scores") +
    theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold", colour = colour),
    plot.subtitle = element_text(hjust = 0.5, size = 12, face = "bold", colour = colour),
    panel.grid = element_blank()) +
  ggeasy::easy_remove_legend() +
  ggeasy::easy_x_axis_labels_size(size = 9)
```

@@

```{r cog-scores}
#| message: false
#| fig-width: 12
#| label: fig-cognitive-test-scores
#| fig-cap: Cognitive test scores (2016 - 2022)

@@

@fig-dementia-proportions shows the proportion of dementia classifications from HRS 2016 - 2022. The variable **Missing** represents the procrastination HRS respondents who were not yet interviewed by the HRS. Since the analysis focuses on participants included in the 2020 HRS wave, any retroactive analysis of prior waves may result in missing data for certain individuals

```{r cog-class}
#| fig-width: 12
#| label: fig-dementia-proportions
#| fig-cap: Proportion of dementia classifications per year

@@

```{r cog-transition}
#| fig-width: 12
#| label: fig-dementia-transitions
#| fig-cap: Dementia transitions (2016 - 2022)

@@

```{r cog-transition-absorbing}
#| fig-width: 12
#| label: fig-dementia-transitions-absorbing
#| fig-cap: Dementia transitions with dementia as an absorbing state (2016 - 2022)

>>>>

# Setting up

```{r}
#| label: set-up
#| code-fold: true
#| code-summary: "Check out my code"
#| warning: false
#| message: false

# Packages ---------------------------------------------------------------------
pacman::p_load(
  dplyr,         # For data manipulation
  tidyr,         # For even more data manipulation
  stringr,       # String manipulation
  ggplot2,       # For cool plots
  ggalluvial     # For alluvial diagrams
)

# Functions --------------------------------------------------------------------
func_files <- c(
  here::here("R/count_transitions.R"),
  here::here("R/extract-years.R"),
  here::here("R/plot_cognitive_scores.R"),
  here::here("R/plot_transitions.R")
)

purrr::walk(func_files, source)

# Theme ------------------------------------------------------------------------
colour = "#212427"

theme_set(
  theme_minimal() +
    theme(
      plot.title = element_text(
        hjust = 0.5,
        size = 14,
        face = "bold",
        colour = colour
      ),
      plot.subtitle = element_text(
        hjust = 0.5,
        size = 10,
        colour = colour
      ),
      axis.title = element_text(
        size = 10,
        face = "bold",
        colour = colour
      ),
      strip.text = element_text(
        size = 10,
        face = "bold",
        colour = colour
      ),
      panel.grid = element_blank(),
    )
)
```

## Reading in data

```{r}
#| label: data

data <- read.csv(here::here("analysis/data/data.csv")) 
```

@@

```{r}
#| label: count-transitions

dementia_data <- data |>
  count_transitions(years = seq(1996, 2022, by = 2))

@@

```{r}
#| fig-width: 8
#| code-fold: true
#| code-summary: "Check out my code"
#| label: fig-missing-data
#| fig-cap: Occurance of missing data across HRS waves (participants were gathered from the 2020 HRS wave [in red])
#| warning: false

@@

data |>
  select(any_of(paste0("cogtot27_imp", 1996:2022))) |>
  rename_with(~ str_replace(., "cogtot27_imp", ""), starts_with("cogtot27")) |>
  visdat::vis_dat() +
  labs(title = "Missing data across HRS Waves (1996 - 2022)", 
       subtitle = "Total Cognitive Scores") +
  theme(
    plot.title = element_text(
      hjust = 0.5, size = 14, face = "bold", colour = colour),
    plot.subtitle = element_text(
      hjust = 0.5, size = 12, face = "bold", colour = colour),
    panel.grid = element_blank(),
    axis.text.x = element_text(size = 9, colour = colours),
    legend.position = "none"
  )

@@

```{r}
#| fig-width: 8
#| code-fold: true
#| code-summary: "Check out my code"
#| label: fig-complete-data
#| fig-cap: HRS dataset after processing
#| warning: false

data |>
  select(any_of(paste0("cogtot27_imp", 2016:2022))) |>
  rename_with(~ str_replace(., "cogtot27_imp", ""), starts_with("cogtot27")) |>
  na.omit() |>
  visdat::vis_dat() +
  labs(title = "Complete dataset across HRS Waves (2016 - 2022)", 
       subtitle = "Total Cognitive Scores") +
  theme(
    plot.title = element_text(
      hjust = 0.5, size = 14, face = "bold", colour = colour),
    plot.subtitle = element_text(
      hjust = 0.5, size = 12, face = "bold", colour = colour),
    axis.text.x = element_text(size = 9),
    panel.grid = element_blank(),
    legend.position = "none"
  )
```

@@

```{r}
#| message: false
#| fig-width: 12
#| label: fig-cognitive-test-scores
#| fig-cap: Cognitive test scores (2016 - 2022)

@@

@fig-dementia-proportions shows the proportion of dementia classifications from HRS 2016 - 2022.

```{r}
#| fig-width: 12
#| label: fig-dementia-proportions
#| fig-cap: Proportion of dementia classifications per year

@@

```{r}
#| fig-width: 12
#| label: fig-dementia-transitions
#| fig-cap: Dementia transitions (2016 - 2022)

@@

```{r}
#| fig-width: 12
#| label: fig-dementia-transitions-absorbing
#| fig-cap: Dementia transitions with dementia as an absorbing state (2016 - 2022)

### 2025-07-02 analysis/data-overview.qmd
Updating analysis scripts

---
title: "Data Overview"
format: 
  html:
    code-tools: false
---

@@

- **Normal Cognition:** Individuals in this category show no significant signs of cognitive impairment and are able to function independently in daily activities.
- **Mild Cognitive Impairment (MCI):** This stage reflects a slight but noticeable decline in cognitive abilities, such as memory or thinking skills, that goes beyond what would be expected with normal aging but does not yet interfere significantly with daily life.
- **Dementia:** This classification is marked by more severe cognitive deficits, impacting memory, reasoning, and the ability to perform everyday tasks. Dementia encompasses various neurodegenerative conditions, with Alzheimer's disease being the most common.

@@

For previous waves of HRS data (1995 - 2020) there is a researcher contributed dataset of dementia classifications [@langa2023]. Researchers have used this dataset to study the trajectories of cognitive aging, dementia risk, and related health outcomes in older adults. However, with the recent release of the 2022 HRS data, **these classifications have yet to be updated**

@@

In order to address this we developed the `LWC2022` package [@Monaghan2024] to replicate the methodology of @langa2023. This package automates the classification process, ensuring consistency with previous waves and providing researchers with an up-to-date resource for studying cognitive decline and dementia across all available HRS waves. @fig-lwc illustrates the LWC workflow.

![LWC Workflow](../files/workflow.svg){#fig-lwc}

>>>>

---
title: "Data Overview"
bibliography: ../manuscript/references.bib
csl: ../manuscript/csl/apa.csl
---

@@

- **Normative Cognitive Function:** Individuals in this category show no significant signs of cognitive impairment and are able to function independently in daily activities.
- **Mild Cognitive Impairment (MCI):** This stage reflects a slight but noticeable decline in cognitive abilities, such as memory or thinking skills, that goes beyond what would be expected with normal aging but does not yet interfere significantly with daily life.
- **Dementia:** This classification is marked by more severe cognitive deficits, impacting memory, reasoning, and the ability to perform everyday tasks. Dementia encompasses various neurodegenerative conditions, with Alzheimer's disease being the most common.

@@

For previous waves of HRS data (1995 - 2020) there is a researcher contributed dataset of dementia classifications [@Langa2023]. Researchers have used this dataset to study the trajectories of cognitive aging, dementia risk, and related health outcomes in older adults. However, with the recent release of the 2022 HRS data, **these classifications have yet to be updated**

@@

In order to address this we developed the `LWC2022` package [@Monaghan2024b] to replicate the methodology of @Langa2023. This package automates the classification process, ensuring consistency with previous waves and providing researchers with an up-to-date resource for studying cognitive decline and dementia across all available HRS waves.

### 2025-07-02 analysis/hypothesis.qmd
Launching site
Updating analysis scripts

---
title: "Hypotheses"
format: 
  html:
    code-tools: false
---

## Primary Hypothesis

@@

- **H1:** Higher levels of procrastination are associated with:
  - Increased risk of transitioning from Normal Cognition to MCI $(\beta = x, p < .05)$
  - Increased risk of transitioning from MCI to Dementia $(\beta = y, p < .05)$

- **H0:** There is no association between procrastination and cognitive decline transitions

In order to test this hypothesis we will be using a **discrete time Markov Model**

>>>>

---
title: "Hypotheses"
format: 
  html:
    code-tools: false
    toc: false
---

## Hypothesis 1

@@

- **H0:** There are no group differences in procrastination scores between cognitive groups (normative cognitive function; MCI; Dementia).
- **H1:** There are group differences in procrastination scores between cognitive groups (normative cognitive function; MCI; Dementia)

## Hypothesis 2

- **H0:** There is no association between procrastination and cognitive decline transitions
- **H1:** Higher scores on a measure of procrastination are associated with:
  - Increased risk of transitioning from normative cognitive function to MCI / dementia $(\beta = x, p < .05)$
  - Increased risk of transitioning from MCI to Dementia $(\beta = y, p < .05)$

In order to test our hypotheses we will be using both a **non-parametric ANOVA** and a post-hoc test to compare procrastination scores between cognitive groups, and a **discrete time Markov Model** to assess the association between procrastination and cognitive decline transitions.
